---
title: "Forth Day Multi Omics"
author: "Mark davids"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
# setting global options for the rmd markdown
knitr::opts_chunk$set(echo = T) # include chunck code in html
knitr::opts_chunk$set(cache = F) # use chuck results from cache if not modified (speed)
knitr::opts_chunk$set(warning = F) # ignore warnings in html
knitr::opts_chunk$set(message = F) # ignore messages in html
knitr::opts_chunk$set(fig.width = 12) # Figure width
```

```{r}
library(vegan)
library(phyloseq)
library(ggplot2)
library(microbiome)
library(ggpubr)
library(tidyverse)
library(mixOmics)
```

# Multi Omics

Here we will explore two methods for "multiomics" analysis.
The dataset used are baseline samples from the MPS study, which is composed of healthy FMT donors and the metSyn patients. The two omics datasets are fecal microbiome composition and fasted plasma metabolomics.

During the first analysis we will test if distance matrix of each of the two datasets have similar structures. The second analysis consists of a Regularized Canonical Correlation Analysis with which we will try to identify between datasets relation ships.

# Load and subset to equal datasets

```{r}
MPS.metabs <- readRDS("MPS.metabs.RDS")
taxa_names(MPS.metabs) <- MPS.metabs@tax_table[,"BIOCHEMICAL"]

MPS.16S <- readRDS("MPS.16S.triads.mod.RDS")
MPS.16S <- prune_samples(MPS.16S@sam_data$Sample_Type!="PostFMT", MPS.16S)
MPS.16S@sam_data$Subject_ID %in% sample_names(MPS.metabs)
sample_names(MPS.16S) <- MPS.16S@sam_data$Subject_ID
taxa_names(MPS.16S) <- make.unique(paste(MPS.16S@tax_table[,6]))


```

# Reorder phyloseq so they are in the same order

```{r}
sample_names(MPS.16S) == sample_names(MPS.metabs)
MPS.metabs <- phyloseq(MPS.metabs@otu_table[, sample_names(MPS.16S)],
                       tax_table(MPS.metabs))

sample_names(MPS.16S) == sample_names(MPS.metabs)
sample_data(MPS.metabs) <- sample_data(MPS.16S)
```

In order to reduce dimensions a bit we will aggregate microbiome data on Genus level. We are also not interested in Xenobiotics from the metabolomics panel.

```{r}
MPS.16S <- microbiome::aggregate_taxa(MPS.16S, "Genus")
MPS.metabs <- prune_taxa(!MPS.metabs@tax_table[,"SUPER PATHWAY"] %in% "Xenobiotics" , MPS.metabs)
```

# Procrustus

In order to run procrustus we will need ordinations of the dataset.
For 16S we use default Bray-Curtis distance.Since metabolite data needs to be scaled we will use clr transformation. After CLR transformation euclidean distance is a very appriopriate distance metrics.
For both sets we will use PCoA for ordination.

To test the likelyhood of getting the obtained fit, we will performa the protest permutation test.

```{r}
ord.16S <- ordinate(MPS.16S, "PCoA", "bray")
ord.metabs <- ordinate(microbiome::transform(MPS.metabs, "clr"), "PCoA", "euclidean")

plot(procrustes(X = ord.16S$vectors, Y = ord.metabs$vectors, scale=T))
protest(X = ord.16S$vectors, Y = ord.metabs$vectors, permutations = 99999)
```

These results show a significant relation between microbiome composition and circulating metabolites. 

# rCCA

CCA is dimensionality reduction approach by maximizing the correlation of the features between the datasets. Variance in the data that is not coherent between both datasets is thereby reduced.

We will apply a regularized form of CCA, which is required if you want to perform CCA on datasets with a higher number of features than samples.
We will reduce the data somewhat by only selecting 100 features with the highest variability.

```{r}
nfeat=100
asvs <- apply(microbiome::transform(MPS.16S, "Z")@otu_table@.Data,1,function(x) sum(abs(x))) %>% sort(decreasing = T) %>% head(nfeat) %>% names()

MPS.16S.reduced <- prune_taxa(asvs, MPS.16S)

metabs <- apply(microbiome::transform(MPS.metabs, "Z")@otu_table@.Data,1,function(x) sum(abs(x))) %>% sort(decreasing = T) %>% head(nfeat) %>% names()

MPS.metabs.reduced <- prune_taxa(metabs, MPS.metabs)

X <- t(MPS.16S.reduced@otu_table@.Data)
Y <- t(MPS.metabs.reduced@otu_table@.Data)
```

# perform the CCA

There are two regularisation methods available.
Here we apply the shrinkage method as feature counts are substantially larger than the number of samples.

```{r}
rcc.MPS <- rcc(X,Y, method = 'shrinkage') 
# examine the optimal lambda values after shrinkage 
plot(rcc.MPS, type = "barplot", main = "Shrinkage") 
```

The scree plot shows that there is a small drop off after two components indicating these two components capture more variance than by chance.

We can plot the samples in this CCA ordination.

```{r}
plotIndiv(rcc.MPS, comp = 1:2, 
          ind.names = NULL,
          group = MPS.16S@sam_data$Sample_Type, rep.space = "XY-variate", 
          legend = TRUE, title = 'MPS, rCCA shrinkage XY-space')
```

This plot shows that if we focus only on coherent variance between the dataset, out healthy donor samples separate from preFMT samples.  
This implies there is a relation between fecal microbiome, plasma metabolites and health status (or other confounding factor)

We can plot the projection of both datasets, similar to the procrustus plot.

```{r}
plotArrow(rcc.MPS, group = MPS.16S@sam_data$Sample_Type, 
          col.per.group = color.mixo(1:5),
          title = '(b) Nutrimouse, shrinkage method')
```
Besides the relations between the samples, we can also use CCA to find correlations between the datasets from the feature perspective.
We can use either a circle plot, or a heatmap to visualize the correlations between the feature sets.

```{r}
plotVar(rcc.MPS, var.names = c(TRUE, TRUE),
        cex = c(3, 3), cutoff = 0.5,
        title = '(b) Nutrimouse, rCCA shrinkage comp 1 - 2')
```

```{r}
cim(rcc.MPS, comp = 1:2, xlab = "metabolites", ylab = "taxa")
```

A difference between groups was not obvious from the unconstrained ordinations

```{r}
plot_ordination(physeq = MPS.16S.reduced, ordination = ordinate(MPS.16S, "PCoA", "bray"), color="Sample_Type")
plot_ordination(physeq = MPS.metabs.reduced, ordination = ordinate(MPS.metabs.reduced, "PCoA", "bray"), color="Sample_Type")
```